
# Getting Started with R Lesson 1
### by Dmitry Grapov



### GOALS:
### * [getting started](#basics) 
### * [basic operators](#operators)
### * [plotting](#plotting)
### * [creating data summaries](#summary)

### For the following examples we will focus on the bare bones to help us learn basic concepts


<a name="basics"/>
### Getting started
```{r,message=FALSE,warning=FALSE}
# this is a comment R doesn't interpret this
10 # works
# a # remove comment and try to type a
```

R knows what ```r 10 ``` is, it is a number which has been defined in the R environment. However a has not been defined and R throws an error when trying to print it to the console.

```{r,message=FALSE,warning=FALSE}
"a" # works
a<-10 # assignment
b<-a+1 # using a pre assigned variable
```

When we give ```r "a" ``` we define that this object is as a character vector "a" which R understands how to print to screen. We can assign ```r a ``` to something R knows about like for instance ```r 10 ```. Having done this we can now use ```r a ``` as a variable to do whatever we want.

<a name="operators"/>
### Arithmetic operators
```{r,message=FALSE,warning=FALSE}
a + b #addition
a - b #subtraction
a * b #multiplication
a / b #division
a^b #exponentiation
a %/% b #integer division
a %% b #modulo (remainder) b%%a makes more sense
```

### Relational operators
```{r,message=FALSE,warning=FALSE}
a==b #is a equal to b
a>b # a is greater than b, use < for less than
a>=b # a is greater than or equal to
```

### Logical operators
```{r,message=FALSE,warning=FALSE}
! #not
a!=b # is a equal to be? could also do !a==b    
#& #and
# | #or
# && #sequential and
# || #sequential or
```

### Lets create a new variable to test the logical operators
```{r,message=FALSE,warning=FALSE}
x<-1:10 # note ':' creates a sequence
x!=a # notice how the shorter a is recycled
(id<-!x>=a) # notice we assigned the results to id and use () to print this
x>a|x<b # check if any criteria specified is TRUE
x>a&x<b # check if all criteria specified is TRUE
```

### Indexing
```{r,message=FALSE,warning=FALSE}
a[1] # get the first 'unit' of a
a[2] #  trying to reference something which doesn't exist, NA stands for missing value
x[2] # this works because x has length >=2
x[c(1,2,6:8)] # can use c() to get many specific elements
x[id] # we can also use a logical to get our object
```

### Object properties
```{r,message=FALSE,warning=FALSE}
str(x) #structure, x is an integer vector (one dimension)
class(id) # is logical
length(x) # length
```

### Next lets load some real data to experiment with
```{r,message=FALSE,warning=FALSE}
data(iris) # we use a function named data to load the iris data
str(iris) # structure
#notice we have 2 dimensions now, rows and columns
# now subset the object as object[rows,columns]
iris[1:5,3:4]
species<-iris$Species # in data.frames and list we can also reference columns with '$'
species<-iris[,"Species"] # or by name
```

### Lets make some plots
```{r,message=FALSE,warning=FALSE}
plot(iris[,1:2]) # plot the first 2 columns
help(plot) #see what other arguments plot could take
plot(iris[,1],species) 
plot(iris[,1]~species) # plot is different because species is factor and formula notation "~" is used
plot(iris[,1]~species, col =c("red","green","blue")) # add a color
plot(iris[,1]~species, col =c("red","green","blue"),ylab=colnames(iris)[1]) # and label, use function colnames() to get column names and take the first columns name
plot(iris) # because we gave the whole data frame R calls pairs() and creates a scatterplot matrix
color<-c("red","green","blue")[species] # we can use the factor to subset our colors to create a color for each point
plot(iris, pch=21,bg=color, main="My Awesome Plot!") # here we give custom point shape 'pch', border 'col' and inner color 'bg' as well as a title 'main'
```

### Next lets experiment with getting summary statics
```{r,message=FALSE,warning=FALSE}
mean(iris) # we want to get the mean but giving the whole data.frame with the factor does not make sense?
mean(iris[,1]) # this works
mean(iris[,"Species"]) # this was the issue, it is not numeric (we could coerce 'as.numeric' but why?)
mean(iris[iris$Species=="setosa",1]) # here we subset the rows to only get values for the species setosa and return the mean for column 1 for this group
```

#### Next lets get all column means for the species setosa
```{r,message=FALSE,warning=FALSE}
apply(iris[iris$Species=="setosa",1:4],2,mean) #here we use function apply to for each column  use the function mean (see help(apply)) 
# we also removed species because we know mean wont work on this and also causes bad behaviour for other columns
```

#### Now lets get the means for each species
```{r,message=FALSE,warning=FALSE}
big.l<-split(iris[,1:4], iris$Species) # create a list holding a data.frame for each level of the species
(res<-lapply(big.l,apply,2,mean)) # get means for each species and variable
data.frame(res) # get results combined
t(data.frame(res)) # transpose results 
```

#### Using this basic syntax we can exucute any custom function that we want (more to come later on R libraries to this this in an elegant and efficient way).
#### Lets define a custom function to return the formatted mean and standard deviation, and then use it in the code above.
```{r,message=FALSE,warning=FALSE}
mean.sd<-function(x){
  paste0(round(c(mean(x),sd(x)),1),collapse=" +/- ") # our custom function
  
  }
t(sapply(big.l,apply,2,mean.sd))
```









